Comparison of clusters to cell types for `r params$library_id`

### 2024-12-05 analyses/cell-type-ewings/template_notebooks/clustering-workflow/02-clustering-celltypes.Rmd
update report to use a single resolution and algorithm

---
title: "Comparison of clusters to cell types for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: "SCPCL000822"
  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
  singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
  cluster_results_file: "../../results/cluster_exploration/SCPCS000490/SCPCL000822_clusters.tsv"
  marker_genes_file: "../../references/visser-all-marker-genes.tsv"
---

TODO: Update this notebook to take as input a clustering algorithm, nn (range of nn?), resolution, and optional objective function to display in SingleR and marker gene plots

- The clusters are then compared to the results from running `SingleR` in the `aucell-singler-annotation.sh` workflow.
- We then look at marker gene expression across the clusters and assign each cluster to a cell type.

@@

```{r}
# check that sce and results files exist
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file),
  "SingleR results file does not exist" = file.exists(params$singler_results_file),
  "Marker genes file does not exist" = file.exists(params$marker_genes_file)
)
```

@@

# source in helper functions: calculate_sum_markers()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)

@@

# read in singler results
singler_classification_df <- readr::read_tsv(params$singler_results_file) |>
  dplyr::rename("cell_id" = "barcodes")
```

## Clusters vs. `SingleR` {.tabset}

Now we will compare the clustering results to the cell type assignments obtained from `SingleR` in the `aucell-singler-annotation.sh` workflow. 
To compare results we will calculate the Jaccard similarity index between clusters and cell types. 
We expect that good clustering will line up with the cell type assignments so that there is ~ 1 cell type per cluster.

For the plots, we will only display the top 7 cell types identified by `SingleR` and all other cells will be lumped together into `All remaining cell types`.

```{r}
cluster_classification_df <- all_cluster_results |>
  # add in classifications from singler
  dplyr::left_join(singler_classification_df, by = "cell_id") |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )

cluster_classification_df |> 
  split(cluster_classification_df$resolution) |> 
  purrr::map(\(df){cluster_celltype_heatmap(df)})

```

```{r, fig.height=10, fig.width=7}
# get heatmap for showing cluster assignments vs. singler assignments
cluster_celltype_heatmap(cluster_classification_df)
```

## Marker gene expression {.tabset}

@@

# join sum expression columns with clustering results
cluster_exp_df <- all_cluster_results |>
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

```{r, fig.height=7, fig.width=7}
# plot to look at marker gene expression for all cell types across all clusters and all params
k_params <- unique(cluster_exp_df$nn_param) |>
  as.character()

k_params |>
  purrr::map(\(k) {
    plot_marker_genes(cluster_exp_df,
      k_value = k
    )
  })
```

>>>>

---
title: "Comparison of clusters to cell types for `r params$library_id`"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: "hide"
params:
  library_id: "SCPCL000822"
  sce_file: "../../../../data/current/SCPCP000015/SCPCS000490/SCPCL000822_processed.rds"
  singler_results_file: "../../results/aucell_singler_annotation/SCPCS000490/SCPCL000822_singler-classifications.tsv"
  cluster_results_file: "../../results/clustering/SCPCS000490/SCPCL000822_cluster-results.tsv"
  marker_genes_file: "../../references/visser-all-marker-genes.tsv"
  algorithm: "leiden"
  objective_function: "modularity"
  resolution: 1
---

This notebook will compare cluster assignments to cell type assignments and look at expression of tumor and normal marker genes. 
We expect that "good" clusters will be compromised of cells belonging to the same cell type and express marker genes for a single cell type (e.g., only tumor or only immune markers).

Clustering was calculated using the following parameters:

- Algorithm: `r params$algorithm` 
- Objective function: `r params$objective_function`
- Resolution: `r params$resolution`

Results will be shown for all values of nearest neighbors used to perform clustering found in the cluster results file.

@@

```{r}
# check that sce and results files exist
stopifnot(
  "sce file does not exist" = file.exists(params$sce_file),
  "SingleR results file does not exist" = file.exists(params$singler_results_file),
  "Clustering results file does not exist" = file.exists(params$cluster_results_file),
  "Marker genes file does not exist" = file.exists(params$marker_genes_file)
)
```

@@

# source in helper functions: calculate_sum_markers() and plot_faceted_umap()
validation_functions <- file.path(module_base, "scripts", "utils", "tumor-validation-helpers.R")
source(validation_functions)

@@

# read in singler results
singler_classification_df <- readr::read_tsv(params$singler_results_file)

# read in clustering results 
clusters_df <- readr::read_tsv(params$cluster_results_file) |> 
  # filter to specified parameters 
  dplyr::filter(algorithm == params$algorithm,
                objective_function == params$objective_function,
                resolution == params$resolution) |> 
  # modify the nn column to be a character for easy plot labels
  dplyr::mutate(cluster = as.factor(cluster),
                nn_char = glue::glue("nn-{nn}") |> 
                  # make sure 5 comes first 
                  forcats::fct_relevel("nn-5", after = 0)) |> 
  # rename barcode column for joining
  dplyr::rename(barcodes = "cell_id")
```

## UMAP of cluster assignments

```{r}
# get umap embeddings and combine into a data frame with cluster assignments
umap_df <- sce |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  # replace UMAP.1 with UMAP1 and get rid of excess columns
  dplyr::select(barcodes, UMAP1 = UMAP.1, UMAP2 = UMAP.2) |>
  dplyr::left_join(clusters_df, by = "barcodes")
```

```{r fig.height=10, fig.width=10}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster)) +
  geom_point(alpha = 0.5, size = 0.1) +
  facet_wrap(vars(nn),
             labeller = labeller(nn = ~ glue::glue("{.}-nn"))) +
  theme(
    aspect.ratio = 1,
    legend.position = "none", 
    panel.border = element_rect(color = "black", fill = NA)
  )
```

## Clusters vs. `SingleR`

Now we will compare the clustering results to the cell type assignments obtained from `SingleR` in the `aucell-singler-annotation.sh` workflow. 
For the plots, we will only display the top 7 cell types identified by `SingleR` and all other cells will be lumped together into `All remaining cell types`.

```{r}
# prep classifications and clusters for plots 
cluster_classification_df <- umap_df |>
  # add in classifications from singler
  dplyr::left_join(singler_classification_df, by = "barcodes") |>
  dplyr::mutate(
    # first grab anything that is tumor and label it tumor
    # NA should be unknown
    singler_annotation = dplyr::case_when(
      stringr::str_detect(singler_annotation, "tumor") ~ "tumor",
      is.na(singler_annotation) ~ "unknown", # make sure to separate out unknown labels
      .default = singler_annotation
    ) |>
      forcats::fct_relevel("tumor", after = 0),
    # get the top cell types for plotting later
    singler_lumped = singler_annotation |>
      forcats::fct_lump_n(7, other_level = "All remaining cell types", ties.method = "first") |>
      forcats::fct_infreq() |>
      forcats::fct_relevel("All remaining cell types", after = Inf)
  )
```

First, we just look at the cell type assignments on the UMAP.

```{r fig.height=10, fig.width=10}
celltype_plot_df <- cluster_classification_df |> 
  dplyr::select(barcodes, UMAP1, UMAP2, singler_lumped)

plot_faceted_umap(celltype_plot_df, singler_lumped)
```

To compare results we will calculate the Jaccard similarity index between clusters and cell types. 
We expect that good clustering will line up with the cell type assignments so that there is ~ 1 cell type per cluster.

```{r, fig.height=15, fig.width=7}
# get heatmap for showing cluster assignments vs. singler assignments
cluster_celltype_heatmap(cluster_classification_df)
```

## Marker gene expression

@@

# join sum expression columns with clustering results
cluster_exp_df <- clusters_df |>
  dplyr::left_join(gene_exp_df, by = "barcodes")
```

```{r, fig.height=7, fig.width=7}
# plot to look at marker gene expression for all cell types across all clusters and all params
k_params <- unique(cluster_exp_df$nn_char)

k_params |>
  purrr::map(\(k) {
    k_num <- stringr::str_remove(k, "nn-")

    plot_marker_genes(cluster_exp_df,
      k_value = k_num
    )
  })
```
